Interpretable Deepfake Voice Detection: A Hybrid Deep-Learning Model and Explanation Evaluation

### 2025-03-05 draft.tex
small changes

There has been considerable attention on identifying audio deepfakes using classical and sophisticated deep learning models~\cite{wang2020deepsonar,becker2024audiomnist,muller_does_2022,yang_robust_2024,wang_asvspoof_2020}. Methods for fake voice identification can be broadly categorized into two classes: methods with handcrafted feature extraction and methods with end-to-end deepfake detectors~\cite{dixit2023review,li2025survey}. In the former, detection approaches first extract various features from the voices. These high-dimensional feature values are then passed through complex deep learning models to determine whether the voice is real or synthesized (i.e., fake). End-to-end fake voice identification methods generally optimize the feature extraction process and classification task jointly. Both methods, however, have shown promise in the field to produce reliable results, as we will discuss in section \ref{sec:related}.

@@

The high-level buidling-blocks of our hybrid interpretable deep fake voice identification model are presented in Fig.~\ref{fig:overview}.

\subsection{Feature Extraction}
The extracted features from audio samples can be categorise in two different class: perceptible and imperceptible features. The perceptible features are features that can be perceived by the human ear, often vocal qualities such as jitter, shimmer or pitch fluctuation that have a wide range of uses even outside of audio classification such as diagnosis of disease~\cite{chaiwongyen_deepfake-speech_2023}, while imperceptible features are typically out of the range of human hearing, and may not directly reflect a vocal quality, otherwise referred to as ``speaker-independent'' features~\cite{liu_hidden--wave_2023}. In order to increase the likelihood that the explanations are useful and understandable to the end user, we focused on using multiple perceptible features as input to the classifier. However, we used two imperceptible features with the hypothesis, based on previous research~\cite{barrington_single_2023,chaiwongyen_contribution_2022,chaiwongyen_deepfake-speech_2023,li_comparative_2022}, that they would positively increase model performance. We extracted the following perceptible features:

@@

\subsection{Hybrid Fake Voice Detection Model}

@@

% REWRITE
In order to address the second problem, we needed to, for each input feature, figure out what the output of the respective sub-model would be. This would allow us to transform the input into the one-dimensional input for the terminus model from the given sample in the input layers in every sub-model. After the processing of the sub-model, we catch the representation before the processing of the terminus model, which are one-dimensional similar representations of the features. This allows us to use the explainer to find a weight of each input of the terminus model in relation to the end result, prediction. We generate the sub-model results for a random sample of the training set to use as a reference for the localized model approximations from the LIME explainer.
% -------
% We approached the second problem by generating intermediate data for the LIME-evaluation, the prediction outputs of the sub-models, in order to determine what the input of the terminus model would be. We generate intermediate data by first decomposing the surrogate model into its component models, the sub-models and the terminus, and for each sub-model, running a prediction on the given input features, so that we have the intermediate results of each sub-model before the processing of the terminus model. We stored the results of the predictions of the sub-models and used them for further analysis. This was done not only for the sample explanation data but also for a random sampling of the training data, so that the LIME explainer could use this data to generate localized estimated of model behaviour.

@@

\subsubsection{Evaluation of the Explanations}
In order to make an assessment the potential of this method of generating explanations, we needed a way to measure the influence of each feature on the average explanation generated using this method. In other words, we needed a method to assess the aggregated results of many explanations. Other methods such as SHAP (SHapley Additive exPlanations) \cite{lundberg2017unified} offer ways to explain features on a global scale, but due to their differing methodology in comparison with LIME, we see aggregated evaluations of a large number of local explanations as the best way to represent and assess the usefulness of our method. However, due to the lack of existing metrics for this purpose, we propose two new metrics with the intent of aggregate evaluation features present within local explanations.
% In order to assess if the explanations have the potential to be useful to an end user, we pursued the goal of summarizing many explanations generated using the previously described method together in a meaningful way in order to make a conclusion about the general usefulness of the individual features and the influence and potential use of the perceptible features. However, due to a lack of existing standardized metrics for the aggregate, global-scale evaluation of local explanations, we propose two metrics, with which we can better assess the usefulness of our method.

In order to contextualize the metrics, we will make the following definitions.
    \begin{itemize}
        \item Let \(S\) be the set of samples with \(s \in S\) as an element of the set.
        \item Let \(F\) be the set of features with \(f \in F\) as an element of the set.
        \item Let \(w_{fs}\) be the weight value of feature \(f\) of sample \(s\) as
            produced by the explainer.
        \item Let \(l_{s}\) be the correct label of the sample \(s\).
    \end{itemize}
The first metric we propose is the mean of the absolute values of the weights of each feature within a set of explanations.
% The first metric we will define is the mean of the absolute values of the weights of the explanations, summarized along the sample axis, delivering one value per feature. 
We will further refer to this metric as importance, and we define it mathematically as follows for every feature \(f \in F\) extracted from sample \(s \in S\):
\begin{equation}
    Imp(f, s) = \frac{\sum_{s \in S} |w_{fs}|}{|S|}.
\end{equation}

@@

%Using the methods described in the previous section, we performed experiments training and testing the surrogate model as well as generating many explanations for the purpose of aggregate evaluation. In this section, we will first discuss the dataset that we used for our experiments, the results and performance of the surrogate model, followed by an evaluation of the aggregated explanations.
\subsection{Dataset}
We trained and tested out model employing \textit{In-the-Wild} dataset. This dataset focuses on the generalization of audio deepfake detection models by collecting real-world data, in comparison to other research that used more controlled laboratory conditions~\cite{muller_does_2022}. We have chosen it for these experiments because of the aforementioned focus on generalization, its relative recency compared to some others and the fact that it has also been used previously by some other experiments, which provides a useful perspective against which we can compare our method.

@@

\subsection{Results}
We present experimental results for all variations of our proposed model in Table~\ref{table:eval-results}. The table details the following information: The type of terminus model used, where ``perceptron'' does not include hidden layers, MPL(3) is a multi-perceptron layer model with three hidden layers and CNN (3) is a convolutional neural network with three hidden layers; we used different set of features, where ``standard" is the feature set as described above including only one pitch-fluctuation feature. However, with expanded pitch fluctuations, we combine the same feature set with the addition of more pitch fluctuation features with different comparison distances; the training batch size and the number of epochs per batch if there were more than one; followed by several performance metrics.

@@

\if false
		\begin{figure}[htbp]
			\centering
			\includegraphics[width=0.7\textwidth]{images/roc_cterm.png}
			\caption{ROC curve of the best-performing model, with 3 hidden layers in the terminus
			and that standard feature set.}
			\label{fig:roc_cterm}
		\end{figure}
		\begin{figure}[htbp]
			\centering
			\includegraphics[width=0.7\textwidth]{images/roc_mpf.png}
			\caption{ROC curve of the model with expanded pitch fluctuation features.}
			\label{fig:roc_mpf}
		\end{figure}
		\begin{figure}[htbp]
			\centering
			\includegraphics[width=0.7\textwidth]{images/roc_mpf_cterm.png}
			\caption{ROC curve of the model with expanded pitch fluctuation features and a
			terminus with three hidden layers.}
			\label{fig:roc_mpf_cterm}
		\end{figure}
\fi

@@

%The model that resulted in the best accuracy and EER had the standard feature set and was paired with a terminus with three hidden perceptron layers. We trained this model on the first 10000 samples of the dataset, and we evaluated it on the rest of the dataset using the methods previously described. It demonstrated an evaluation accuracy of 96.3\% with the default threshold of 0.5, as well as an EER of approximately 0.03214. The best AUC score, however, was achieved in another experiment, where the model had expended pitch fluctuations and a terminus model without hidden layers, with a value of 0.92727.

Figure~\ref{ROC_curves} depicts the ROC curves for three of the models, which were selected for use in explanation generation in the following section.

@@

\subsection{Evaluating explanations generated by LIME}
In order to obtain a comprehensive understanding of the quality of the explanations generated by LIME, we selected the three best-performing models and generated explanations for the first 500 samples of the testing data. These three models achieved the best performance in terms of Accuracy and Equal Error Rate (EER) or AUC. They include: a model using the standard feature set with a terminus model containing three hidden layers, model with expanded pitch-fluctuation features and a perceptron terminus, which achieved the best AUC result, and model with expanded pitch-fluctuation features and a terminus containing three hidden layers.

@@

\subsubsection{Model with Expanded Pitch-Fluctuation Features and MLP Terminus}
Table~\ref{table:exp-results-mpf-cterm} and Fig.~\ref{fig:exp_mpf_cterm} present the results for the evaluation of the explanations for the model with pitch fluctuation features and MLP terminus. The model combining multiple pitch-fluctuation features with a terminus containing multiple hidden layers performed similarly to the other two models. However, in this case, the pitch fluctuations had no influence on the outcomes. Unlike the previous examples, the intensities feature had a negative trust value. Nevertheless, the trust values for the jitter and shimmer features remained positive, reinforcing their reliability as perceptible features for generating meaningful explanations.
		\begin{figure}
			\centering
			\includegraphics[width=0.7\textwidth]{images/exp_mpf_cterm.png}
			\caption{Graphical representation of the evaluation of generated explanations in terms of \textit{trust} and \textit{importance} for the model with expanded pitch fluctuation features and a terminus with MLP.}
			\label{fig:exp_mpf_cterm}
		\end{figure}
%In order to obtain a full picture of the quality of the explanations generated by LIME, we have selected three best performing models and generated explanations on the first 500 samples of the testing data. Those three models achieved best performance in terms of Accuracy and EER, and applies the standard features and with three hidden layers in the terminus model, expanded pitch-fluctuation features and perceptron terminus, which achieved the best AUC result, as well as the expanded pitch-fluctuation features and a terminus with three hidden layers, respectively. We selected these models were selected because we aimed at evaluating the best performing models, as well as observing how the changed pitch-fluctuation feature affected the evaluation of the explanations. We then aggregated and summarized the explanations generated using both of the previously introduced metrics. A summary of the evaluation in terms of our proposed metric \textit{trust} and \textit{importance} score summarized by the feature is presented in tables~\ref{table:exp-results-cterm}, \ref{table:exp-results-more-pitch-flucs} and \ref{table:exp-results-mpf-cterm}. The same statistics are also represented graphically in the figures \ref{fig:exp_cterm}, \ref{fig:exp_mpf} and \ref{fig:exp_mpf_cterm} respectively.

%\subsubsection{Model with Standard Feature Set and Complex Terminus}
%Measuring by both metrics, the MFCCs had the most positive and the most correct influence on the classification results of the model, with the jitter features as a distant second for trust and importance. This indicates that the MFCCs had not only the most influence, but also the most accurate influence on the result of the classification, while the other features had either little, no or negative influence on the classification. However, jitter features are still notable for being perceptible features with a positive trust value.

%\subsubsection{Model with Expanded Pitch-Fluctuation Features}
%This model had comparable results, with the main difference being that all metrics have non-zero values, meaning that all the present features had an influence on the classifications which were used to generate the explanations. As with the previous model, HNRs and mel-spectrograms had negative overall trust values, with the addition of the onset strengths also presenting a negative value, implying that these features have caused more harm than good in the tested classifications. In this example, both jitter and shimmer features are notable for their positive trust values indicating their potential as trustworthy perceptible features for explanations.

%\subsubsection{Model with Expanded Pitch-Fluctuation Features and Complex Terminus}
%The model with multiple pitch-fluctuation features in combination with a terminus containing multiple hidden layers fared similarly to the other two, this time with the pitch fluctuations having no influence on the outcome whatsoever. Unlike the previous examples, the trust value of the intensities feature was negative here. The trust of the jitter and shimmer features, however, remain positive.

@@

\subsection{Viability and Potential of the Explanations}
%Because of these models' inclusion of several perceptible features, we made the hypothesis that this method has the potential to produce explanations that can be used as the basis for results more presentable to an end user.

>>>>

There has been considerable attention on identifying audio deepfakes using classical and sophisticated deep learning models~\cite{wang2020deepsonar,becker2024audiomnist,muller_does_2022,yang_robust_2024,wang_asvspoof_2020}. Methods for fake voice identification can be broadly categorized into two classes: methods with handcrafted feature extraction and methods with end-to-end deepfake detectors~\cite{dixit2023review,li2025survey}. In the former, detection approaches first extract various features from the voices. These high-dimensional feature values are then passed through complex deep learning models to determine whether the voice is real or synthesized (i.e., fake). End-to-end fake voice identification methods generally optimize the feature extraction process and classification task jointly. Both methods, however, have shown promise in the field to produce reliable results.

@@

The high-level buidling-blocks of our hybrid interpretable deep fake voice identification model are presented in Fig.~\ref{fig:overview}. The architecture is more thoroughly discussed in Section \ref{sec:method_arch}.

\subsection{Feature Extraction}
The extracted features from audio samples can be categorized into two different classes: perceptible and imperceptible features. The perceptible features are features that can be perceived by the human ear, often vocal qualities such as jitter, shimmer or pitch fluctuation that have a wide range of uses even outside of audio classification such as diagnosis of disease~\cite{chaiwongyen_deepfake-speech_2023}, while imperceptible features are typically out of the range of human hearing, and may not directly reflect a vocal quality, otherwise referred to as ``speaker-independent'' features~\cite{liu_hidden--wave_2023}. In order to increase the likelihood that the explanations are useful and understandable to the end user, we focused on using multiple perceptible features as input to the classifier. However, we used two imperceptible features with the hypothesis, based on previous research~\cite{barrington_single_2023,chaiwongyen_contribution_2022,chaiwongyen_deepfake-speech_2023,li_comparative_2022}, that they would positively increase model performance. We extracted the following perceptible features:

@@

\subsection{Hybrid Fake Voice Detection Model} \label{sec:method_arch}

@@

In order to address the second problem, we needed to, for each input feature, figure out what the output of the respective sub-model would be. This would allow us to transform the input into the one-dimensional input for the terminus model from the given sample in the input layers in every sub-model. After the processing of the sub-model, we catch the representation before the processing of the terminus model, which are one-dimensional similar representations of the features. This allows us to use the explainer to find a weight of each input of the terminus model in relation to the end result, prediction. We generate the sub-model results for a random sample of the training set to use as a reference for the localized model approximations from the LIME explainer.

@@

\subsubsection{Evaluation of the Explanations}
In order to make an assessment the potential of this method of generating explanations, we needed a way to measure the influence of each feature on the average explanation generated using this method. In other words, we needed a method to assess the aggregated results of many explanations. Other methods such as SHAP (SHapley Additive exPlanations) \cite{lundberg2017unified} offer ways to explain features on a global scale, but due to their differing methodology in comparison with LIME, we see aggregated evaluations of a large number of local explanations as the best way to represent and assess the usefulness of our method. However, due to the lack of existing metrics for this purpose, we propose two new metrics with the intent of aggregate evaluation features present within local explanations.

In order to contextualize the metrics, we will make the following definitions.
    \begin{itemize}
        \item Let \(S\) be the set of samples with \(s \in S\) as an element of the set.
        \item Let \(F\) be the set of features with \(f \in F\) as an element of the set.
        \item Let \(w_{fs}\) be the weight value of feature \(f\) of sample \(s\) as
            produced by the explainer.
        \item Let \(l_{s}\) be the correct label of the sample \(s\).
    \end{itemize}
The first metric we propose is the mean of the absolute values of the weights of each feature within a set of explanations.
We will further refer to this metric as importance, and we define it mathematically as follows for every feature \(f \in F\) extracted from sample \(s \in S\):
\begin{equation}
    Imp(f, s) = \frac{\sum_{s \in S} |w_{fs}|}{|S|}.
\end{equation}

@@

\subsection{Dataset}
We trained and tested out model employing \textit{In-the-Wild} dataset. This dataset focuses on the generalization of audio deepfake detection models by collecting real-world data, in comparison to other research that used more controlled laboratory conditions~\cite{muller_does_2022}. We have chosen it for these experiments because of the aforementioned focus on generalization, its relative recency compared to some others and the fact that it has also been used previously by some other experiments, which provides a useful perspective against which we can compare our method.

@@

\subsection{Results}
We present experimental results for all variations of our proposed model in Table~\ref{table:eval-results}. The table details the following information: The type of terminus model used, where ``perceptron'' does not include hidden layers, MPL(3) is a multi-perceptron layer model with three hidden layers and CNN (3) is a convolutional neural network with three hidden layers; we used different set of features, where ``standard'' is the feature set as described above including only one pitch-fluctuation feature. However, with expanded pitch fluctuations, we combine the same feature set with the addition of more pitch fluctuation features with different comparison distances; the training batch size and the number of epochs per batch if there were more than one; followed by several performance metrics.

@@

@@

Figure~\ref{ROC_curves} depicts the ROC curves for three of the models, which were selected for use in explanation generation in Section \ref{sec:method_metrics}.

@@

\subsection{Evaluating explanations generated by LIME} \label{sec:method_metrics}
In order to obtain a comprehensive understanding of the quality of the explanations generated by LIME, we selected the three best-performing models and generated explanations for the first 500 samples of the testing data. These three models achieved the best performance in terms of Accuracy and Equal Error Rate (EER) or AUC. They include: a model using the standard feature set with a terminus model containing three hidden layers, model with expanded pitch-fluctuation features and a perceptron terminus, which achieved the best AUC result, and model with expanded pitch-fluctuation features and a terminus containing three hidden layers.

@@

\subsubsection{Model with Expanded Pitch-Fluctuation Features and MLP Terminus}
Table~\ref{table:exp-results-mpf-cterm} and Fig.~\ref{fig:exp_mpf_cterm} present the results for the evaluation of the explanations for the model with pitch fluctuation features and MLP terminus. The model combining multiple pitch-fluctuation features with a terminus containing multiple hidden layers performed similarly to the other two models. However, in this case, the pitch fluctuations had no influence on the outcomes. Unlike the previous examples, the intensities feature had a negative trust value. Nevertheless, the trust values for the jitter and shimmer features remained positive, reinforcing their reliability as perceptible features for generating meaningful explanations.
		\begin{figure}
			\centering
			\includegraphics[width=0.7\textwidth]{images/exp_mpf_cterm.png}
			\caption{Graphical representation of the evaluation of generated explanations in terms of \textit{trust} and \textit{importance} for the model with expanded pitch fluctuation features and a terminus with MLP.}
			\label{fig:exp_mpf_cterm}
		\end{figure}

@@

\subsection{Viability and Potential of the Explanations}
